Forecasting the United States 2024 Presidential Election

### 2024-11-04 paper/paper.qmd
added details

For states lacking recent polling data, we incorporated historical averages of Democratic and Republican support from elections since 2000. This decision is based on the assumption that certain states are very unlikely to change their voting patterns, reducing the necessity for current polling data. The ten states that do not have polling data [REF] have not changed parties in at least the last 20 plus years. This may be why they do not have polling data, as it would be a waste of resources for agencies to poll them. Alternatively, imputation methods or political regional averages could have been used, but historical averages are used for their simplicity and relevance.

@@

Comprehensive diagnostics were performed to assess model convergence, fit, and predictive performance. Graphs and more details can be found in the relevant appendices. We evaluated model convergence using the R-hat statistic and trace plots. All R-hat values for model parameters were below 1.1, indicating satisfactory convergence. Trace plots revealed well-mixed chains without discernible trends, further supporting the reliability of the MCMC sampling process [REF].

@@

Residual plots, depicting residuals versus fitted values, showed notable clumping in the central region for both models, along with vertical lines. This pattern may indicate unmodeled heterogeneity or data limitations, suggesting areas where the model's fit could be improved. Despite these observations, the overall residual distribution did not reveal significant systematic errors, supporting the model's general adequacy.

@@

Our model predicts that Harris has a 68.8% chance of winning the US presidential election. While Trump only has a 30.1% chance of winning. There is also a 1.1% chance of a tie On average across the 1,000 simulations Harris wins 279.21 Electoral Votes, while Trump wins on average 258.79 votes.

@@

#| echo: false
#| warning: false
#| message: false
election_results <- data.frame(
  Metric = c("Chance of Winning", "Average Electoral Votes"),
  Harris = c("68.8%", "279.21"),
  Trump = c("30.1%", "258.79"),
  stringsAsFactors = FALSE
)

@@

@@

Consequently, our forecast suggests that candidates should concentrate their efforts on swing states, as increased campaigning in these areas could significantly impact voter turnout and the popular vote. Additionally, enlisting high-profile figures or celebrities, like Elon Musk or LeBron James—who have recently endorsed Trump and Harris, respectively—could enhance their appeal in these swing states. These endorsements would help candidates reach diverse and niche demographics, broadening their voter base and potentially swaying undecided voters.

@@

Advantages

Disadvantages and drawbacks

>>>>

For states lacking recent polling data, we incorporated historical averages of Democratic and Republican support from elections since 2000. This data was collected from a github repository [REF] who collected their data from Wikipedia. This decision is based on the assumption that certain states are very unlikely to change their voting patterns, reducing the necessity for current polling data. The ten states that do not have polling data [REF] have not changed parties in at least the last 20 plus years. This may be why they do not have polling data, as it would be a waste of resources for agencies to poll them. Alternatively, imputation methods or political regional averages could have been used, but historical averages are used for their simplicity and relevance.

@@

Comprehensive diagnostics were performed to assess model convergence, fit, and predictive performance. Graphs and more details can be found in the relevant appendices. We evaluated model convergence using the R-hat statistic and trace plots. Trace plots revealed well-mixed chains without discernible trends, further supporting the reliability of the MCMC sampling process [REF].

@@

Residual plots, depicting residuals versus fitted values, showed notable clumping in the central region for both models, along with vertical lines[REF]. This pattern may indicate unmodeled heterogeneity or data limitations, suggesting areas where the model's fit could be improved. Despite these observations, the overall residual distribution did not reveal significant systematic errors, supporting the model's general adequacy.

@@

Our model predicts that Harris has a 67.4% chance of winning the US presidential election. While Trump only has a 31.7% chance of winning. There is also a 0.9% chance of a tie On average across the 1,000 simulations Harris wins 279.33 Electoral Votes, while Trump wins on average 258.67 votes.

@@

#| echo: false
#| warning: false
#| message: false
election_results <- data.frame(
  Metric = c("Chance of Winning", "Average Electoral Votes"),
  Harris = c("67.4%", "279.33"),
  Trump = c("31.7%", "258.67"),
  stringsAsFactors = FALSE
)

@@

Our model does not take into account state correlations based on past election data and research based insights into voter preference. For example a model developed by Chernov, Elenev, and Song, posits if Harris were to win Pennsylvania, then her chances of winning Nevada would increase from 24% to 68% [https://anderson-review.ucla.edu/wp-content/uploads/2024/10/voter_preferences_updated.pdf]. These correlations between states could have a large impact on election predictions and our model is limited by not modelling these interactions.

@@

Consequently, our forecast suggests that candidates should concentrate their efforts on swing states, as increased campaigning in these areas could significantly impact voter turnout and the popular vote. Additionally, enlisting high-profile figures or celebrities, like Elon Musk or LeBron James—who have recently endorsed Trump and Harris, respectively—could enhance their appeal in these swing states. These endorsements would help candidates reach diverse and niche demographics, broadening teheir voter base and potentially swaying undecided voters.

@@

Some key advantages of these markets is that they are real time, so the market adjusts quickly to important and material news. Where as poll are sluggish to incorporate big changes as new polls need to be sent out and analysed. Another advantage is that these markets may attract sophisticated investors that could employ methods that differ from polling but still offer predictive capability. The larger the trade volume the more likely these actors are to be present and wash out any unsophisticated trading.

On the other hand the main disadvantage is that all of the investors in this market essentially completely anonymous. This means there is no way of checking whether a large investor, who has significant price setting power, is indeed sophisticated or not. Furthermore, because these markets have only recently become popular or even legal in some cases one might expect these markets to be full of amateurs. These markets may display a bais as well, where the largest market Polymarket [REF] is based on crypto-currency, and Trump is popular among crypto-enthusiasts, and large financial institutions are underrepresented in the crypto-market.

This is a new frontier in forecasting and including data from these markets may prove beneficial. Or these markets may be nothing more than a hobby, and hold no useful data.
